
---
title: "NOAA Data Preparation"
author: "Steven Lio, Patricia Luo"
date: "1/29/2022"
output: html_document
---

```{r setup, include=FALSE}
library(rnoaa)
library(ggplot2)
library(plotly)

library(tidyverse)
library(broom)
library(lubridate)
library(forecast)

knitr::opts_chunk$set(echo = TRUE)
```

## Weather Data Preparation

Procedure to prepare NOAA weather data.

## Function to fetch station data
```{r}
get_station_data <- function (stationid) {
  ghcnd_search(stationid = stationid, var = "all", 
               date_min = "1950-01-01", date_max = Sys.Date())
}
```


## Function to fetch all data given station id
```{r}
get_data <- function(station_ids,verbose=FALSE) {
  weather_data <- tibble(id=as.character(),
                         location=as.character(),
                         var=as.character(),
                         value=as.numeric(),
                         date=as.Date(NA),
                         mflag=as.character(),
                         qflag=as.character(),
                         sflag=as.character()
                         )
  
  for (s in 1:nrow(station_id)){
    if (verbose) {print(paste0("Fetching data for location: ",station_id[s,]$location))}
    a <- get_station_data(station_id[s,]$id)
    
    for (i in 1:length(a)){
      if (verbose) {
        print(paste0("Variable: ",colnames(a[[i]])[2]))
        print(paste0("Rows: ",nrow(a[[i]])))
      }
      
      weather_data <- weather_data %>%
        add_row(id = a[[i]]$id,
                location = rep(station_id[s,]$location,nrow(a[[i]])),
                var = colnames(a[[i]])[2],
                value = a[[i]][,colnames(a[[i]])[2]] %>% pull(),
                date = a[[i]]$date,
                mflag = a[[i]]$mflag,
                qflag = a[[i]]$qflag,
                sflag = a[[i]]$sflag)
      if (verbose) {
        print("")
      }
    }
  }
  
  weather_data
}
```

## List of station ID to be fetch
```{r}
station_id <- tibble(location=c("Washington DC","Liestal","Kyoto","Vancouver"),
                     id = c("USC00186350","GME00127786","JA000047759","CA001108395"))
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
weather_data <- get_data(station_id)
```

## Data restructure and expand data to include missing dates.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
var_impute <- unique(weather_data %>% select(id, var))
weather_data_complete <- weather_data[0,]

for (i in 1:nrow(var_impute)){
  print(paste0("Complete series for ",var_impute[i,1] %>% pull(),", variable: ",var_impute[i,2]))
  ts_raw <- weather_data %>% 
    filter(id==var_impute[i,1] %>% pull(),
           var==var_impute[i,2] %>% pull())
  
  ts_complete_row <-  ts_raw %>%
    complete(date = seq.Date(min(date), max(date), by="day")) %>%
    mutate(id=ifelse(is.na(id),unique(na.omit(id)),id),
           location=ifelse(is.na(location),unique(na.omit(location)),location),
           var=ifelse(is.na(var),unique(na.omit(var)),var),
           value=ifelse(var %in% c("tmax","tmin","prcp","tobs","wesd","tavg","wsfg"),value/10,value),
           mflag = ifelse(is.na(mflag)," ",mflag),
           qflag = ifelse(is.na(qflag)," ",qflag),
           sflag = ifelse(is.na(sflag)," ",sflag)
    )
  
  weather_data_complete <- weather_data_complete %>%
    add_row(ts_complete_row)
  
  new_row <- weather_data_complete %>% 
    filter(id==var_impute[i,1] %>% pull(),
           var==var_impute[i,2] %>% pull()) %>%
    nrow()
  
  print(paste0("Original: ",nrow(ts_raw)," rows. New: ",new_row))
}

weather_data_complete_summary <- weather_data_complete %>%
  group_by(id, location, var) %>%
  summarise(dt.min = min(date),
            dt.max = max(date),
            missing_dt = n() - (max(date)-min(date)+1),
            cnt = n(),
            val.na = sum(is.na(value)),
            val.min = min(na.omit(value)),
            val.max = max(na.omit(value)),
            val.avg = mean(na.omit(value)),
            val.sd = sd(na.omit(value))
            )

if (weather_data_complete_summary %>% filter(missing_dt > 0) %>% nrow() > 0){
  print("Still have missing data")
}
```
## Plot for `tmax` by locations
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=7, fig.height=8}
plt_data <- weather_data_complete %>% filter(var == "tmax" & date >= "2010-01-01") %>% mutate(location=factor(location))
plt_data%>%
  group_by(location) %>%
  do(p=plot_ly(., x = ~date, y = ~value, color = ~location, type = "scatter",mode="lines")) %>%
  subplot(nrows = 4, shareX = TRUE, shareY = TRUE)
```

## Imputation and Prediction function for `tmax` using Linear Regression (with sine and cosine terms) and ARIMA.
```{r}
ma<-function(x, window=5, incl=TRUE, method="center"){
  ma_x <- c()
  
  for (i in 1:length(x)){
    if (method=="center"){
      ma_x[i]<-mean(x[max(1,i-window):min(length(x),i+window)],na.rm=TRUE)
      ma_x[i]<-ifelse(is.na(ma_x[i]),NA,ma_x[i])
    } else if (method=="forward") {
      ma_x[i]<-mean(x[max(1,j-window+1):j],na.rm=TRUE)
      ma_x[i]<-ifelse(is.na(ma_x[i]),NA,ma_x[i])
    } else if (method=="backward") {
      ma_x[i]<-mean(x[length(x):1][max(1,j-window+1):j],na.rm=TRUE)
      ma_x[i]<-ifelse(is.na(ma_x[i]),NA,ma_x[i])
    }
  }
  return(ma_x)
}

var_predict <- function(df, end_date="2031-12-31", max_freq=50,biasweight=0.9) {
  
  #df <- tmax_df
  #end_date <- "2031-12-31"
  #max_freq = 10
  
  var_df <- df %>% 
    mutate(value_impute = ifelse(is.na(df$value),lag(df$value, 365),df$value))
  var_df <- df %>% 
    mutate(value_impute = ifelse(is.na(df$value_impute),lead(df$value_impute, 365),df$value_impute))
  
  pred_date <- seq.Date(from=min(var_df$date),to=as.Date(end_date), by = "day")
  day_num <- 1:length(pred_date) 
  doy<-yday(pred_date)
  
  var_pred <- tibble(id = unique(na.omit(var_df$id)),
                      location = unique(na.omit(var_df$location)),
                      var = unique(na.omit(var_df$var)),
                      date = pred_date,
                      day_num = day_num,
                      doy = doy,
                      value = c(var_df$value,rep(NA,as.Date(end_date) - max(var_df$date))),
                      value_impute = c(var_df$value_impute,rep(NA,as.Date(end_date) - max(var_df$date))))
  
  #LM
  print("Running Linear Regression")
  lm_df <- var_pred %>% 
              select(date, day_num,doy, value_impute) %>%
              na.omit()
  
  ssp<-spectrum(lm_df$value_impute,plot=FALSE)
  n_per <- min(max_freq,length(ssp$freq))
  per <- 1/ssp$freq[order(-ssp$spec)][1:n_per]
  sincos <-""
  for (i in 1:n_per){
    sincos<-paste(sincos, "+sin(2*pi/",per[i],"*day_num)+cos(2*pi/",per[i],"*day_num)+sin(2*pi/",per[i],"*doy)+cos(2*pi/",per[i],"*doy)",sep="")
  }
  Linear <- lm(as.formula(paste0("value_impute~",sincos)),data=lm_df)
  
  var_pred$LM.pred <- predict(Linear,var_pred)

  var_pred$LM.predbias <- var_pred %>% 
  mutate(resid=value_impute-LM.pred) %>%
  group_by(doy) %>%
  mutate(bias = sample(na.omit(resid),1),
         LM.predbias=LM.pred + bias*biasweight) %>%
  ungroup() %>% select(LM.predbias) %>% pull()
  
  #print("Running ARIMA")
  #ARIMA
  #var_arima <- stlf(ts(var_df$value_impute,start=min(var_df$date),frequency=365),
                     #h=(as.Date(end_date) - max(var_df$date)),method="arima")
  
  #bias <-sample(var_df$value_impute-var_arima$fitted, length(var_arima$mean))
  #var_pred$ARIMA.pred <- c(var_arima$fitted, var_arima$mean)
  #var_pred$ARIMA.predbias <- c(var_arima$fitted, var_arima$mean + bias*0.8)
  
  return(var_pred)
}
```



## Predict tmax for Washington DC

```{r}
tmax_df <- weather_data_complete %>%
  filter(var == "tmax" & location == "Washington DC") %>%
  mutate(day_num = 1:n(),
         doy = yday(date),
         na.flag=is.na(value),
         value_impute = value
         )

tmax_pred <- var_predict(tmax_df, end_date="2031-12-31", max_freq=150, biasweight=0.8)

weather_data_complete_washington <- bind_rows(
weather_data_complete %>% 
    filter(location=="Washington DC") %>%
    select(id, location, date, var, value)
,
tmax_pred %>% select(id,location,date,value_impute,LM.pred,LM.predbias) %>%
  pivot_longer(cols=c(value_impute,LM.pred,LM.predbias),names_to="var",values_to="value")) %>%
                      
    pivot_wider(names_from=var, values_from = value) %>% 
  mutate(year=year(date),
           qtr=paste0("Q",quarter(date)),
           month=month(date),
           day_num = 1:n(),
           doy = yday(date),
         tmax_impute=value_impute,
           tmax_LM_pred=LM.pred,
         tmax_LM_predbias=LM.predbias)

fig <- plot_ly(data=tmax_pred,x=~date,y=~value,type="scatter",mode="lines",name="original")
fig <- fig %>% add_trace(data=tmax_pred,x=~date,y=~ma(value_impute,window=2),type="scatter",mode="lines",name="MA")
fig <- fig %>% add_trace(data=tmax_pred,x=~date,y=~LM.pred,type="scatter",mode="lines",name="LM")
fig <- fig %>% add_trace(data=tmax_pred,x=~date,y=~LM.predbias,type="scatter",mode="lines",name="LMBias")
fig


sqrt(mean((tmax_pred$value-tmax_pred$LM.pred)^2,na.rm=TRUE))
sqrt(mean((tmax_pred$value-tmax_pred$LM.predbias)^2,na.rm=TRUE))
sqrt(mean((ma(tmax_pred$value_impute,window=7)-tmax_pred$LM.pred)^2,na.rm=TRUE))
sqrt(mean((ma(tmax_pred$value_impute,window=7)-tmax_pred$LM.predbias)^2,na.rm=TRUE))


fig<-plot_ly(data=tmax_pred,x=~value_impute,type="histogram",histnorm="probability",name="ori")
fig <- fig %>% add_trace(data=tmax_pred,x=~LM.pred,type="histogram",histnorm="probability",name="LM")
fig <- fig %>% add_trace(data=tmax_pred,x=~LM.predbias,type="histogram",histnorm="probability",name="LMbias")
fig
```

## Predict tmax for Liestal

```{r}
tmax_df <- weather_data_complete %>%
  filter(var == "tmax" & location == "Liestal") %>%
  mutate(day_num = 1:n(),
         doy = yday(date),
         na.flag=is.na(value),
         value_impute = value
         )

tmax_pred <- var_predict(tmax_df, end_date="2031-12-31", max_freq=150, biasweight=0.8)

weather_data_complete_liestal <- bind_rows(
weather_data_complete %>% 
    filter(location=="Liestal") %>%
    select(id, location, date, var, value)
,
tmax_pred %>% select(id,location,date,value_impute,LM.pred,LM.predbias) %>%
  pivot_longer(cols=c(value_impute,LM.pred,LM.predbias),names_to="var",values_to="value")) %>%
                      
    pivot_wider(names_from=var, values_from = value) %>% 
  mutate(year=year(date),
           qtr=paste0("Q",quarter(date)),
           month=month(date),
           day_num = 1:n(),
           doy = yday(date),
         tmax_impute=value_impute,
           tmax_LM_pred=LM.pred,
         tmax_LM_predbias=LM.predbias)

fig <- plot_ly(data=tmax_pred,x=~date,y=~value,type="scatter",mode="lines",name="original")
fig <- fig %>% add_trace(data=tmax_pred,x=~date,y=~ma(value_impute,window=2),type="scatter",mode="lines",name="MA")
fig <- fig %>% add_trace(data=tmax_pred,x=~date,y=~LM.pred,type="scatter",mode="lines",name="LM")
fig <- fig %>% add_trace(data=tmax_pred,x=~date,y=~LM.predbias,type="scatter",mode="lines",name="LMBias")
fig


sqrt(mean((tmax_pred$value-tmax_pred$LM.pred)^2,na.rm=TRUE))
sqrt(mean((tmax_pred$value-tmax_pred$LM.predbias)^2,na.rm=TRUE))
sqrt(mean((ma(tmax_pred$value_impute,window=7)-tmax_pred$LM.pred)^2,na.rm=TRUE))
sqrt(mean((ma(tmax_pred$value_impute,window=7)-tmax_pred$LM.predbias)^2,na.rm=TRUE))


fig<-plot_ly(data=tmax_pred,x=~value_impute,type="histogram",histnorm="probability",name="ori")
fig <- fig %>% add_trace(data=tmax_pred,x=~LM.pred,type="histogram",histnorm="probability",name="LM")
fig <- fig %>% add_trace(data=tmax_pred,x=~LM.predbias,type="histogram",histnorm="probability",name="LMbias")
fig
```

## Predict tmax for Kyoto

```{r}
tmax_df <- weather_data_complete %>%
  filter(var == "tmax" & location == "Kyoto") %>%
  mutate(day_num = 1:n(),
         doy = yday(date),
         na.flag=is.na(value),
         value_impute = value
         )

tmax_pred <- var_predict(tmax_df, end_date="2031-12-31", max_freq=150, biasweight=0.8)

weather_data_complete_kyoto <- bind_rows(
weather_data_complete %>% 
    filter(location=="Kyoto") %>%
    select(id, location, date, var, value)
,
tmax_pred %>% select(id,location,date,value_impute,LM.pred,LM.predbias) %>%
  pivot_longer(cols=c(value_impute,LM.pred,LM.predbias),names_to="var",values_to="value")) %>%
                      
    pivot_wider(names_from=var, values_from = value) %>% 
  mutate(year=year(date),
           qtr=paste0("Q",quarter(date)),
           month=month(date),
           day_num = 1:n(),
           doy = yday(date),
         tmax_impute=value_impute,
           tmax_LM_pred=LM.pred,
         tmax_LM_predbias=LM.predbias)

fig <- plot_ly(data=tmax_pred,x=~date,y=~value,type="scatter",mode="lines",name="original")
fig <- fig %>% add_trace(data=tmax_pred,x=~date,y=~ma(value_impute,window=2),type="scatter",mode="lines",name="MA")
fig <- fig %>% add_trace(data=tmax_pred,x=~date,y=~LM.pred,type="scatter",mode="lines",name="LM")
fig <- fig %>% add_trace(data=tmax_pred,x=~date,y=~LM.predbias,type="scatter",mode="lines",name="LMBias")
fig


sqrt(mean((tmax_pred$value-tmax_pred$LM.pred)^2,na.rm=TRUE))
sqrt(mean((tmax_pred$value-tmax_pred$LM.predbias)^2,na.rm=TRUE))
sqrt(mean((ma(tmax_pred$value_impute,window=7)-tmax_pred$LM.pred)^2,na.rm=TRUE))
sqrt(mean((ma(tmax_pred$value_impute,window=7)-tmax_pred$LM.predbias)^2,na.rm=TRUE))


fig<-plot_ly(data=tmax_pred,x=~value_impute,type="histogram",histnorm="probability",name="ori")
fig <- fig %>% add_trace(data=tmax_pred,x=~LM.pred,type="histogram",histnorm="probability",name="LM")
fig <- fig %>% add_trace(data=tmax_pred,x=~LM.predbias,type="histogram",histnorm="probability",name="LMbias")
fig
```

## Predict tmax for Vancouver

```{r}
tmax_df <- weather_data_complete %>%
  filter(var == "tmax" & location == "Vancouver") %>%
  mutate(day_num = 1:n(),
         doy = yday(date),
         na.flag=is.na(value),
         value_impute = value
         )

tmax_pred <- var_predict(tmax_df, end_date="2031-12-31", max_freq=150, biasweight=0.8)

weather_data_complete_vancouver <- bind_rows(
weather_data_complete %>% 
    filter(location=="Vancouver") %>%
    select(id, location, date, var, value)
,
tmax_pred %>% select(id,location,date,value_impute,LM.pred,LM.predbias) %>%
  pivot_longer(cols=c(value_impute,LM.pred,LM.predbias),names_to="var",values_to="value")) %>%
                      
    pivot_wider(names_from=var, values_from = value) %>% 
  mutate(year=year(date),
           qtr=paste0("Q",quarter(date)),
           month=month(date),
           day_num = 1:n(),
           doy = yday(date),
         tmax_impute=value_impute,
           tmax_LM_pred=LM.pred,
         tmax_LM_predbias=LM.predbias)

fig <- plot_ly(data=tmax_pred,x=~date,y=~value,type="scatter",mode="lines",name="original")
fig <- fig %>% add_trace(data=tmax_pred,x=~date,y=~ma(value_impute,window=2),type="scatter",mode="lines",name="MA")
fig <- fig %>% add_trace(data=tmax_pred,x=~date,y=~LM.pred,type="scatter",mode="lines",name="LM")
fig <- fig %>% add_trace(data=tmax_pred,x=~date,y=~LM.predbias,type="scatter",mode="lines",name="LMBias")
fig


sqrt(mean((tmax_pred$value-tmax_pred$LM.pred)^2,na.rm=TRUE))
sqrt(mean((tmax_pred$value-tmax_pred$LM.predbias)^2,na.rm=TRUE))
sqrt(mean((ma(tmax_pred$value_impute,window=7)-tmax_pred$LM.pred)^2,na.rm=TRUE))
sqrt(mean((ma(tmax_pred$value_impute,window=7)-tmax_pred$LM.predbias)^2,na.rm=TRUE))


fig<-plot_ly(data=tmax_pred,x=~value_impute,type="histogram",histnorm="probability",name="ori")
fig <- fig %>% add_trace(data=tmax_pred,x=~LM.pred,type="histogram",histnorm="probability",name="LM")
fig <- fig %>% add_trace(data=tmax_pred,x=~LM.predbias,type="histogram",histnorm="probability",name="LMbias")
fig
```

## Combine results
```{r}
weather_data_complete_full<-bind_rows(weather_data_complete_washington,
weather_data_complete_liestal,
weather_data_complete_kyoto,
weather_data_complete_vancouver) %>%
select(id,location,date,year,qtr,month,day_num,doy,
         tmin,tavg,tmax,tmax_impute,tmax_LM_pred,tmax_LM_predbias,prcp,snow,snwd)
       #,tobs,dapr,mdpr,dasf,mdsf,wesd,wdfg,wsfg,wt01,wt03,wt04,wt05,wt06,wt11)
```

```{r}
loc <- unique(weather_data_complete_full$location)[4]

plt_data <- weather_data_complete_full %>% 
                 filter(location==loc)
fig <- plot_ly(data=plt_data,
               x=~date,
               y=~tmax,
               type="scatter",
               mode="lines",
               name="tmax")
fig <- fig %>% add_trace(data=plt_data,
                         x=~date,
                         y=~tmax_LM_predbias,
                         type="scatter",
                         mode="lines",
                         name="tmax.LM.Pred") %>% layout(title=paste0(loc," (tmax)"))
fig
```


## Output NOAA data
```{r}
write.csv(weather_data_complete_full,"../data/NOAA.csv",row.names = FALSE)
```


```{r}
TRUE
```
